Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges

### 2020-03-09 man/ecosystems/ms_ecosystems.Rmd
change abstract. - incorporate suggestions by DN - remove two last items - include new item (RZ)

* Global change challenges ecosystems in dry locations transformed by intensive human use. Forest resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd. populations in the southern Iberian Peninsula was analyzed in relation to historical records of land use employing dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.

* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood removal) in the current forest structure. Trees were highly sensitive to moisture availability but both primary and secondary growth expressed high resilience to drought events over the short and the long term. Resilience and response to climate followed a water-stress gradient. Primary and secondary growth were generally less resilient to drought in mesic than in xeric sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and wetter) high-elevation stands, but absent in the most xeric (*i.e.* warmer and drier) site.

* Adult oak trees showed high resilience to drought despite being considered relict and at its rear edge. The trees varied in response to the changing climate along a narrow climatic gradient and showed that the ecological and geographical rear edges do not need to meet today in forests with strong land use legacies. Forest resilience including that of regeneration needs to be monitored at xeric of the species under future expected climatic scenarios given that ecological boundaries for species distribution can be narrow and that land-use legacies can play a key role in stand dynamics.

>>>>

* Global change challenges ecosystems in dry locations transformed by intensive human use. Resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd. populations in the southern Iberian Peninsula was analyzed in relation to historical records of land use, employing dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.

* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood removal) in the current forest structure. Trees were highly sensitive to moisture availability but both primary and secondary growth expressed high resilience to drought events over the short and the long term. Resilience and the tree growth response to climate followed a water-stress gradient. A positive growth trend since the late 1970s was particularly evident in mesic (*i.e.* colder and wetter) high-elevation stands, but absent in the most xeric (*i.e.* warmer and drier) site.

* The high values of resilience observed in our study suggest that Q. pyrenaica populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge. These resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies.  This is particularly relevant for rear-edge populations where topographic and biophysical variability can facilitate the existence of refugia
